Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-26 AnalyticStructure_main.tex
update plot and biblio

% Useful packages
\usepackage{graphicx}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{tikz}%,pgfplots}
\usepackage{xifthen}
%\pgfplotsset{compat=1.15}

@@

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm'_\Phi = \dm_{\perp} \doteq \dmnot\dm'_\Phi\op Q_0 + \op Q_0\dm'_\Phi\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p^\Phi(\omega)}\}$, 
that have values in the \emph{unoccupied} subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$%\be
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p}
$. %\ee

@@

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like e.g. a static electric field in the direction $\mathbf F$, $\Phi_s(\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\rho_{\Phi_s}] + \Phi_s$, where we indicate with $\rho_{\Phi_s}$ the GS density of the
perturbed system. We assume that the perturbation strength ($F$ in the above example) is sufficiently small to be consistent with a LR 
treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian 
can be extracted with traditional GS techniques, as by the hypothesis of Eq.~\eqref{SEHelmholtz1},
they would exhibit bound-state behavior.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.56]{CO_alpha+D0_vsBoxSize.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability of the $CO$ with respect to the size of the computational domain.}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.56]{co_spectrum.eps}
\caption{\label{co_spectrum}(Color online) Convergence below threshold of the $CO$ spectrum computed in different computational setup.}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[scale=0.24]{CO_exc.eps}
\caption{(Color online) Excitation spectrum of $CO$. The two curves correspond to two computational domain of different sizes. The inset contains the sole contribution of localized excitations.}
\label{CO_exc}
\end{figure}

\begin{figure}[ht]
\includegraphics[scale=0.24]{C6H6_exc.eps}
\caption{(Color online) Excitation spectrum of $C_6H_6$. The two curves correspond to two computational domain of different sizes. The inset contains the sole contribution of localized excitations.}
\label{C6H6_exc}
\end{figure}

\begin{figure}[ht]
\includegraphics[scale=0.56]{c6h6_excitationLandscape.eps}
\caption{\label{c6h6_excLand}(Color online) Excitation landscape of $C_6H_6$.}
\end{figure}
\section{Criterion of computational reliability for the numerical convergence of the LR}

@@

Express the fluctuation state in terms of excitation using the relation $\ket{f_p(\omega)} = \dm'(\omega)\ket{\psi_p}$ and using formula \eqref{SusceptibilityDef1} to express the excitation
expansion of the response density. Due to this procedure we obtain
\be\lb{FSExcitationDef1}
\ket{f_p(\omega)}   = \left(\sum_a + \int\dd a   \right) \ket{w_p^a} \,
\frac{\bra{\psi_p}\op\Phi(\omega)\ket{w_p^a}}{\omega^2-\Omega_a^2} \;,
\ee 
with $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$. 
Show the convergence of the statical polarizability of the $CO$ in function of the number of excitations. Write formula
\be\lb{AlphaExcitationDef1}
\alpha_{xx} =  \left(\sum_a + \int\dd a   \right) 2 \,
\frac{\left[\trace{\op W_a \op X}\right]^2}{\Omega_a^2} \;,
\ee 
and add the plot that show the convergence rate of the statical polarizability to the reference (finite difference) value in functions of the number of excitations. The plot evidence that the
continuum sector of the excitation spectrum is 
\begin{figure}
\includegraphics[scale=0.56]{CO_alphaX_isto.eps}
\caption{\label{co_AlphaExc}(Color online) Convergence of the statical polarizability of the $CO$ molecule w.r.t the number of excitations.}
\end{figure}

@@

The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The left action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouvnot \op O = \commutator{\hnot}{O} \;,
\ee
and
\be\lb{CouplDef1}
\coupl\op O = \commutator{\op V'[\op O]}{\dmnot}  = \op Q_0\op V'[\op O]\dmnot - \dmnot\op V'[\op O]\op Q_0 \;,
\ee
where $\op V'$ expresses the perturbation induced by the density dependence on ground state Hamiltonian and can be conveniently expressed by defining the 
\emph{scalar coupling kernel}
\be\lb{CouplingKernelDef1}
U\left[\op O; \op O'\right] \equiv  \int \dd \r \dd \r' \trace{\op O \frac{\delta \hnot }{\delta \rho_0(\r,\r')}
} \trace{\op O' \ket{\r} \bra{\r'}} \;, \nn
\ee
on the basis of which
\be\lb{VpDef1}
\op V'[\op O]= 
\int \dd \r \dd \r' U\left[\ketbra{\r}{\r'};\op O\right] \ketbra{\r'}{\r}\;.
\ee
In view of building the spectral decomposition we are interested in finding both the left and right eigenstates of $\Liouv$.   
The right action of $\Liouv$ can be assessed through the equivalence $\trace{\op O'(\Liouv\op O)} = \trace{(\op O'\Liouv)\op O}$, so that:
\be
\op O\Liouv = -\commutator{\hnot}{\op O} + \int \dd\r\dd\r'
U\left[\commutator{\dmnot}{\op O};\ketbra{\r'}{\r}\right]\ketbra{\r}{\r'} \;. \nn
\ee
In the cases relevant for the present analysis the coupling kernel is symmetric with respect to the exchange of its arguments so also the right action of the 
Liouvillian is expressed in terms of $\op V'$, as follows
\begin{align}\lb{LiouvillianRightActionDef1}
\op O\Liouv &= -\commutator{\hnot}{\op O} + V'\left[\commutator{{\dmnot}}{\op O}\right] = -\commutator{\hnot}{\op O} + \nn \\ 
&+ \dmnot\op V'\left[\commutator{{\dmnot}}{\op O}\right]\op Q_0 + \op Q_0\op V'\left[\commutator{\dmnot}{\op O}\right]\dmnot \;.
\end{align}
In the last equivalence we have made usage of the transverse property (see equation \eqref{RhopTransverseDef1}) of the coupling superoperator. Formula 
\eqref{LiouZeroDef1},\eqref{CouplDef1} and \eqref{LiouvillianRightActionDef1} are the starting point for the solution of the eigenvalues equations \eqref{ExcitationOperatorsDef1}.

>>>>

% Useful packages
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{tikz}%,pgfplots}
\usepackage{xifthen}
%\pgfplotsset{compat=1.15}

@@

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm'_\Phi = \dm_{\perp} \doteq \dmnot\dm'_\Phi\op Q_0 + \op Q_0\dm'_\Phi\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p^\Phi(\omega)}\}$, 
that have values in the \emph{unoccupied} subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$%\be
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p}
$. %\ee

@@

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like e.g. a static electric field in the direction $\mathbf F$, $\Phi_s(\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\rho_{\Phi_s}] + \Phi_s$, where we indicate with $\rho_{\Phi_s}$ the GS density of the
perturbed system. We assume that the perturbation strength ($F$ in the above example) is sufficiently small to be consistent with a LR 
treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian 
can be extracted with traditional GS techniques, as by the hypothesis of Eq.~\eqref{SEHelmholtz1},
they would exhibit bound-state behavior.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability (blue continuous line) and of the ground state permanent dipole (red dotted line) 
of the $CO$ molecule with respect to the size of the computational domain. The inset plot evidences the relative percentage convergence using the value associated to the largest 
box as reference.}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.6]{CO_spectrum.eps}
\caption{\label{co_spectrum}(Color online) Convergence below threshold of the $CO$ spectrum computed in different computational setup.}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[scale=0.6]{CO_dos.eps}
\caption{(Color online) Excitation spectrum of $CO$. The two curves correspond to two computational domain of different sizes. The inset contains the sole contribution of localized excitations.}
\label{CO_exc}
\end{figure}

% \begin{figure}[ht]
% \includegraphics[scale=0.6]{C6H6_dos.eps}
% \caption{(Color online) Excitation spectrum of $C_6H_6$. The two curves correspond to two computational domain of different sizes. The inset contains the sole contribution of localized excitations.}
% \label{C6H6_exc}
% \end{figure}
% 
% \begin{figure}[ht]
% \includegraphics[scale=0.56]{C6H6_excLandscape.eps}
% \caption{\label{c6h6_excLand}(Color online) Excitation landscape of $C_6H_6$.}
% \end{figure}

% \begin{figure}[ht]
% \centering
% \begin{subfigure}[]{0.2\textwidth}
% \includegraphics[scale=0.56]{C6H6_excLandscape.eps}
% %\caption{\label{c6h6_excLand}(Color online) Excitation landscape of $C_6H_6$.}
% \end{subfigure}
% \centering
% \begin{subfigure}[b]{0.2\textwidth}
% \includegraphics[scale=0.6]{C6H6_dos.eps}
% \end{subfigure}
% \caption{(Color online) Excitation spectrum of $C_6H_6$. The two curves correspond to two computational domain of different sizes. The inset contains the sole contribution of localized excitations.}
% \end{figure}

\begin{figure}[ht]
\centering
\subfloat[aaaa]
{\label{c6h6_excLand}
\includegraphics[scale=0.56]{C6H6_excLandscape.eps}} \\
\centering
\subfloat[bbbbb]
{\label{C6H6_exc}
\includegraphics[scale=0.6]{C6H6_dos.eps}}
\caption{(Color online) Excitation landscape and DOS of $C_6H_6$.}
\end{figure}

\section{Criterion of computational reliability for the numerical convergence of the LR}

@@

Express the fluctuation state in terms of excitation using the relation $\ket{f_p(\omega)} = \dm'(\omega)\ket{\psi_p}$ and using formula \eqref{SusceptibilityDef1} to express the excitation
expansion of the response density. Due to this procedure we obtain
\be\lb{FSExcitationDef1}
\ket{f_p(\omega)}   = \left(\sum_a + \int\dd a   \right) \ket{w_p^a} \,
\frac{\bra{\psi_p}\op\Phi(\omega)\ket{w_p^a}}{\omega^2-\Omega_a^2} \;,
\ee 
with $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$. 
Show the convergence of the statical polarizability of the $CO$ in function of the number of excitations. Write formula
\be\lb{AlphaExcitationDef1}
\alpha_{xx} =  \left(\sum_a + \int\dd a   \right) 2 \,
\frac{\left[\trace{\op W_a \op X}\right]^2}{\Omega_a^2} \;,
\ee 
and add the plot that show the convergence rate of the statical polarizability to the reference (finite difference) value in functions of the number of excitations. The plot evidence that the
continuum sector of the excitation spectrum is 
\begin{figure}
\includegraphics[scale=0.6]{CO_statPolvsExc.eps}
\caption{\label{co_AlphaExc}(Color online) Convergence of the statical polarizability of the $CO$ molecule w.r.t the number of excitations.}
\end{figure}

@@

The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The left action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouvnot \op O = \commutator{\hnot}{\op O} \;,
\ee
and
\be\lb{CouplDef1}
\coupl\op O = \commutator{\op V'[\op O]}{\dmnot}  = \op Q_0\op V'[\op O]\dmnot - \dmnot\op V'[\op O]\op Q_0 \;,
\ee
where $\op V'$ expresses the perturbation induced by the density dependence on ground state Hamiltonian and can be conveniently expressed by defining the 
\emph{scalar coupling kernel}
\be\lb{CouplingKernelDef1}
U\left[\op O; \op O'\right] \equiv  \int \dd \r \dd \r' \trace{\op O \frac{\delta \hnot }{\delta \rho_0(\r,\r')}
} \trace{\op O' \ket{\r} \bra{\r'}} \;, \nn
\ee
on the basis of which
\be\lb{VpDef1}
\op V'[\op O]= 
\int \dd \r \dd \r' U\left[\ketbra{\r}{\r'};\op O\right] \ketbra{\r'}{\r}\;.
\ee
In view of building the spectral decomposition we are interested in finding both the left and right eigenstates of $\Liouv$.   
The right action of $\Liouv$ can be assessed through the equivalence $\trace{\op O'(\Liouv\op O)} = \trace{(\op O'\Liouv)\op O}$, so that:
\be
\op O\Liouv = -\commutator{\hnot}{\op O} + \int \dd\r\dd\r'
U\left[\commutator{\dmnot}{\op O};\ketbra{\r'}{\r}\right]\ketbra{\r}{\r'} \;. \nn
\ee
In the cases relevant for the present analysis the coupling kernel is symmetric with respect to the exchange of its arguments so also the right action of the 
Liouvillian is expressed in terms of $\op V'$, as follows
\begin{align}\lb{LiouvillianRightActionDef1}
\op O\Liouv &= -\commutator{\hnot}{\op O} + V'\left[\commutator{{\dmnot}}{\op O}\right] = -\commutator{\hnot}{\op O} + \nn \\ 
&+ \dmnot\op V'\left[\commutator{{\dmnot}}{\op O}\right]\op Q_0 + \op Q_0\op V'\left[\commutator{\dmnot}{\op O}\right]\dmnot \;.
\end{align}
In the last equivalence we have made usage of the transverse property (see equation \eqref{RhopTransverseDef1}) of the coupling superoperator. Formula 
\eqref{LiouZeroDef1},\eqref{CouplDef1} and \eqref{LiouvillianRightActionDef1} are the starting point for the solution of the eigenvalues equations \eqref{ExcitationOperatorsDef1}.
